---
title: "RStudio and RMarkdown"
subtitle: "An Introduction to Nice Things in R"
author: "Frances Hung"
date: "10/6/2022"
output:
  html_document:
    toc: true
    toc_float: true
---
echo = showing output

Setup Options chunk:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.height = 12,
                      fig.width = 4,
                      include = TRUE)

```

Welcome to the first series of workshops on Quantitative Methods for HIV Researchers! This first series is an introduction to R that will include basic R constructs, followed by an introduction to R's tools for data cleaning and visualization.

notes: example

# Main heading

## Sub heading

### Sub sub heading

*italics*

**bold**

## Why R Helps with Reproducibility

R is a language widely used by statisticians for data cleaning, visualization, and analysis. Unlike doing analysis in Excel or some other mediums, *all output* is created by and can be traced back to code. 

This nice interface for working with R is called RStudio (soon to be renamed Posit). There are usually four windows open: environment/Git, the file system, the console, and whatever code you're working on. You can either work on the DCC or on a local version of RStudio.

## RMarkdown

RMarkdown is a special type of file that can be created in RStudio. It can integrate R code with plain text to run/document analysis and create reports. Code goes in special sections called chunks, and we can format plain text using markdown commands or the visual tab. We can even put in-line code in plain text. For example, today is `r Sys.Date()`.

The shortcut on windows for a code chunk is "r - Shift -Tab". In the below code chunks, we write some code and R evaluates and prints output.

Open new code chunk:
```{r label}

```


```{r division}
8/4
```



```{r printingHi}
print("Hi! Welcome to RMarkdown.")
2+1
```


